---
title: 'STAT 420: Data Analysis Project - Proposal'
author: "Rafal Tytyk (rtytyk2), Maximilian Trumpf (mtrumpf2)"
date: ''
output:
  html_document:
    theme: flatly
    toc: yes
    fig_width: 10
    fig_height: 5
  pdf_document: default
urlcolor: cyan
---

***

## Project overview

### Project title 

Prediction of student high school performance considering demographic and social features

### Project team

| Name              | Netid    | E-Mail                | Role                                                             |
|-------------------|----------|-----------------------|-------------------------------------------------------------------|
| Rafal Tytyk       | rtytyk2  | rtytyk2@illinois.edu  | No specific role, team will work entire on project collaboratively |
| Maximilian Trumpf | mtrumpf2 | mtrumpf2@illinois.edu | No specific role, team will work entire on project collaboratively |

## Proposed Dataset

### Description

#### Overview

We propose to use the *"Student Performance"* dataset. The dataset documents student performance in secondary education in two Portugese schools. 

The original data provides two datasets covering two distinct subjects: Mathematics (**395 observations**) and Portugese (**649 observations**). Both datasets include the exact same set of attributes (**33 attriutes**, see table below). When merging the two datasets, one will find **382 observations/students** who are present in the data and have taken both courses. (Merged based on ` "school", "sex", "age", "address", "famsize", "Pstatus", "Medu", "Fedu", "Mjob", "Fjob", "reason", "nursery", "internet"`). We are planning to run the analysis primarily on the `Portugese` dataset since it contains more observations than the `Math` dataset (649 observations vs 395 observations). If promising, we will also model the performance in the `Math` dataset and compare the `Portugese` with the `Math` model in the course of the analysis.

The main goal of this project is to predict student performance based on demographic and social features. The dataset includes a variety of predictors. Currently we are considering the following ones as most important response and predictor variables (this list might change during the analysis):

* Main response: **G3 - final grade**
* Main predictors
    * **Sex** (categorial)
    * **Age** (numeric)
    * **Family size** (categorial)
    * **Time spend with various activites** (numeric from 1 to 5; see full list of activites below)
    * **Parents' education** (numeric from 1 to 4)
    
#### Full list of attributes

| Id | Name       | Description                                                                                                                                            |
|----|------------|--------------------------------------------------------------------------------------------------------------------------------------------------------|
| 1  | school     | student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira)                                                                       |
| 2  | sex        | student's sex (binary: 'F' - female or 'M' - male)                                                                                                     |
| 3  | age        | student's age (numeric: from 15 to 22)                                                                                                                 |
| 4  | address    | student's home address type (binary: 'U' - urban or 'R' - rural)                                                                                       |
| 5  | famsize    | family size (binary: 'LE3' - less or equal to 3 or 'GT3' - greater than 3)                                                                             |
| 6  | Pstatus    | parent's cohabitation status (binary: 'T' - living together or 'A' - apart)                                                                            |
| 7  | Medu       | mother's education (numeric: 0 - none, 1 - primary education (4th grade), 2 - 5th to 9th grade, 3 - secondary education or 4 - higher education) |
| 8  | Fedu       | father's education (numeric: 0 - none, 1 - primary education (4th grade), 2 - 5th to 9th grade, 3 - secondary education or 4 - higher education) |
| 9  | Mjob       | mother's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other')                       |
| 10 | Fjob       | father's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other')                       |
| 11 | reason     | reason to choose this school (nominal: close to 'home', school 'reputation', 'course' preference or 'other')                                           |
| 12 | guardian   | student's guardian (nominal: 'mother', 'father' or 'other')                                                                                            |
| 13 | traveltime | home to school travel time (numeric: 1 - 1 hour)                                                                                                       |
| 14 | studytime  | weekly study time (numeric: 1 - 10 hours)                                                                                                              |
| 15 | failures   | number of past class failures (numeric: n if 1<=n<3, else 4)                                                                                           |
| 16 | schoolsup  | extra educational support (binary: yes or no)                                                                                                          |
| 17 | famsup     | family educational support (binary: yes or no)                                                                                                         |
| 18 | paid       | extra paid classes within the course subject (Math or Portuguese) (binary: yes or no)                                                                  |
| 19 | activities | extra-curricular activities (binary: yes or no)                                                                                                        |
| 20 | nursery    | attended nursery school (binary: yes or no)                                                                                                            |
| 21 | higher     | wants to take higher education (binary: yes or no)                                                                                                     |
| 22 | internet   | Internet access at home (binary: yes or no)                                                                                                            |
| 23 | romantic   | with a romantic relationship (binary: yes or no)                                                                                                       |
| 24 | famrel     | quality of family relationships (numeric: from 1 - very bad to 5 - excellent)                                                                          |
| 25 | freetime   | free time after school (numeric: from 1 - very low to 5 - very high)                                                                                   |
| 26 | goout      | going out with friends (numeric: from 1 - very low to 5 - very high)                                                                                   |
| 27 | Dalc       | workday alcohol consumption (numeric: from 1 - very low to 5 - very high)                                                                              |
| 28 | Walc       | weekend alcohol consumption (numeric: from 1 - very low to 5 - very high)                                                                              |
| 29 | health     | current health status (numeric: from 1 - very bad to 5 - very good)                                                                                    |
| 30 | absences   | number of school absences (numeric: from 0 to 93)                                                                                                      |
| 31 | G1         | first period grade (numeric: from 0 to 20)                                                                                                             |
| 32 | G2         | second period grade (numeric: from 0 to 20)                                                                                                            |
| 33 | G3         | final grade (numeric: from 0 to 20)                                                                                                                    |
  
### Background information

The data was collected by using school reports and questionnaries. 

The data was accessed on 2019-07-02 via [UCI Machine Learning Repository/Student Performance](http://archive.ics.uci.edu/ml/datasets/Student+Performance#). 

The dataset was collected by Paulo Cortez, University of Minho, Guimarães, Portugal, [http://www3.dsi.uminho.pt/pcortez](http://www3.dsi.uminho.pt/pcortez).

### Motivation

The main motivation for this analysis is personal: We, the project authors (Maximilian Trumpf and Rafal Tytyk), are both parents ourselves. Being confronted with the situation that our children will go to school soon, we want to give them the best environment for a good performance in school. This analysis will help us to identify the main factors affecting performance in school.

### Evidence data loaded into `R`

First, load the data.
```{r message=FALSE}
students.por = read.csv("data/student-por.csv", sep = ";")
students.mat = read.csv("data/student-mat.csv", sep = ";")
```

Then, print first values of the main response variable `G3`.
```{r}
head(students.por$G3)
```
